 

### 2021-06-22 vertical-thermal-review.Rmd
text edits

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest. 
Leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests. 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the shape of profile modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

>>>>

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d) than open canopy forests such as needle-leaf forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
